---
title: "Octopus_trajectory_analysis"
author: "Enora Geslain"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set the environment and load needed packages

```{r setup, echo=FALSE}
setwd("C:/Users/enora/Documents/work/octopus_SC_Monocle")
library(monocle3) # v. 1.4.25
library(Seurat) # v. 5.3.0
library(SeuratWrappers) # v. 0.4.0
library(anndata) # v. 0.8.0
library(ggplot2) # v. 3.5.2
library(dplyr) # v. 1.1.4
library(tidyverse) # v. 2.0.0
library(clustree) # v. 0.5.1
library(cluster) # v. 2.1.8.1
library(magrittr) # v. 2.0.3
library(patchwork) # v. 1.3.0
```

# Loading raw feature matrices of each batch of the 3 stages and subset them to the cells of the Scanpy UMAP

```{r load_datasets, echo=FALSE}
## Load stage IX batch 1
octo_stage_IX_1 <- Read10X(data.dir = "raw_feature_bc_matrix_stageIX-1")
## Create the Seurat object
octo_stage_IX_1 <- CreateSeuratObject(counts = octo_stage_IX_1[["Gene Expression"]], project = "stage_IX_1")
## Create vector containing cells from Scanpy UMAP
cells_IX_1 <- readLines("cells_IX_1.txt")
## Subset the Seurat object
octo_stage_IX_1_subset <- subset(octo_stage_IX_1, cells = cells_IX_1)
## Remove old Seurat object
remove(octo_stage_IX_1)

## Same for stage IX batch 2
octo_stage_IX_2 <- Read10X(data.dir = "raw_feature_bc_matrix_stageIX-2")
octo_stage_IX_2 <- CreateSeuratObject(counts = octo_stage_IX_2[["Gene Expression"]], project = "stage_IX_2")
cells_IX_2 <- readLines("cells_IX_2.txt")
octo_stage_IX_2_subset <- subset(octo_stage_IX_2, cells = cells_IX_2)
remove(octo_stage_IX_2)

## Same for stage IX batch 3
octo_stage_IX_3 <- Read10X(data.dir = "raw_feature_bc_matrix_stageIX-3")
octo_stage_IX_3 <- CreateSeuratObject(counts = octo_stage_IX_3, project = "stage_IX_3")
cells_IX_3 <- readLines("cells_IX_3.txt")
octo_stage_IX_3_subset <- subset(octo_stage_IX_3, cells = cells_IX_3)
remove(octo_stage_IX_3)

## Same for stage XII batch 1
octo_stage_XII_1 <- Read10X(data.dir = "raw_feature_bc_matrix_stageXII-1")
octo_stage_XII_1 <- CreateSeuratObject(counts = octo_stage_XII_1[["Gene Expression"]], project = "stage_XII_1")
cells_XII_1 <- readLines("cells_XII_1.txt")
octo_stage_XII_1_subset <- subset(octo_stage_XII_1, cells = cells_XII_1)
remove(octo_stage_XII_1)

## Same for stage XII batch 2
octo_stage_XII_2 <- Read10X(data.dir = "raw_feature_bc_matrix_stageXII-2")
octo_stage_XII_2 <- CreateSeuratObject(counts = octo_stage_XII_2[["Gene Expression"]], project = "stage_XII_2")
cells_XII_2 <- readLines("cells_XII_2.txt")
octo_stage_XII_2_subset <- subset(octo_stage_XII_2, cells = cells_XII_2)
remove(octo_stage_XII_2)

## Same for stage XII batch 3
octo_stage_XII_3 <- Read10X(data.dir = "raw_feature_bc_matrix_stageXII-3")
octo_stage_XII_3 <- CreateSeuratObject(counts = octo_stage_XII_3[["Gene Expression"]], project = "stage_XII_3")
cells_XII_3 <- readLines("cells_XII_3.txt")
octo_stage_XII_3_subset <- subset(octo_stage_XII_3, cells = cells_XII_3)
remove(octo_stage_XII_3)

## Same for stage XV batch 1
octo_stage_XV_1 <- Read10X(data.dir = "raw_feature_bc_matrix_stageXV-1")
octo_stage_XV_1 <- CreateSeuratObject(counts = octo_stage_XV_1, project = "stage_XV_1")
cells_XV_1 <- readLines("cells_XV_1.txt")
octo_stage_XV_1_subset <- subset(octo_stage_XV_1, cells = cells_XV_1)
remove(octo_stage_XV_1)

## Same for stage XV batch 2
octo_stage_XV_2 <- Read10X(data.dir = "raw_feature_bc_matrix_stageXV-2")
octo_stage_XV_2 <- CreateSeuratObject(counts = octo_stage_XV_2, project = "stage_XV_2")
cells_XV_2 <- readLines("cells_XV_2.txt")
octo_stage_XV_2_subset <- subset(octo_stage_XV_2, cells = cells_XV_2)
remove(octo_stage_XV_2)

## Clean environment
gc()
```

# Normalize the datasets

```{r normalize, echo=FALSE}
octo_stage_IX_1_subset <- NormalizeData(octo_stage_IX_1_subset, scale.factor = 1000000)
octo_stage_IX_2_subset <- NormalizeData(octo_stage_IX_2_subset, scale.factor = 1000000)
octo_stage_IX_3_subset <- NormalizeData(octo_stage_IX_3_subset, scale.factor = 1000000)
octo_stage_XII_1_subset <- NormalizeData(octo_stage_XII_1_subset, scale.factor = 1000000)
octo_stage_XII_2_subset <- NormalizeData(octo_stage_XII_2_subset, scale.factor = 1000000)
octo_stage_XII_3_subset <- NormalizeData(octo_stage_XII_3_subset, scale.factor = 1000000)
octo_stage_XV_1_subset <- NormalizeData(octo_stage_XV_1_subset, scale.factor = 1000000)
octo_stage_XV_2_subset <- NormalizeData(octo_stage_XV_2_subset, scale.factor = 1000000)
```

# Find the 2000 most variable features

```{r highly_variable, echo=FALSE}
octo_stage_IX_1_subset <- FindVariableFeatures(octo_stage_IX_1_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_IX_2_subset <- FindVariableFeatures(octo_stage_IX_2_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_IX_3_subset <- FindVariableFeatures(octo_stage_IX_3_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_XII_1_subset <- FindVariableFeatures(octo_stage_XII_1_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_XII_2_subset <- FindVariableFeatures(octo_stage_XII_2_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_XII_3_subset <- FindVariableFeatures(octo_stage_XII_3_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_XV_1_subset <- FindVariableFeatures(octo_stage_XV_1_subset, selection.method = "vst", nfeatures = 5000)
octo_stage_XV_2_subset <- FindVariableFeatures(octo_stage_XV_2_subset, selection.method = "vst", nfeatures = 5000)
```

# Integrate the data into one global Seurat object

```{r integrate, echo=FALSE}
## First rename the cell barcodes to make them correspond with the names in annotation file of Scanpy
octo_stage_IX_1_subset <- RenameCells(octo_stage_IX_1_subset, add.cell.id = "IX_1")
octo_stage_IX_2_subset <- RenameCells(octo_stage_IX_2_subset, add.cell.id = "IX_2")
octo_stage_IX_3_subset <- RenameCells(octo_stage_IX_3_subset, add.cell.id = "IX_3")
octo_stage_XII_1_subset <- RenameCells(octo_stage_XII_1_subset, add.cell.id = "XII_1")
octo_stage_XII_2_subset <- RenameCells(octo_stage_XII_2_subset, add.cell.id = "XII_2")
octo_stage_XII_3_subset <- RenameCells(octo_stage_XII_3_subset, add.cell.id = "XII_3")
octo_stage_XV_1_subset <- RenameCells(octo_stage_XV_1_subset, add.cell.id = "XV_1")
octo_stage_XV_2_subset <- RenameCells(octo_stage_XV_2_subset, add.cell.id = "XV_2")

## Create a list with all dataset names
sample.list <- list(octo_stage_IX_1_subset, octo_stage_IX_2_subset, octo_stage_IX_3_subset,
                    octo_stage_XII_1_subset, octo_stage_XII_2_subset, octo_stage_XII_3_subset,
                    octo_stage_XV_1_subset, octo_stage_XV_2_subset)

## Detect features for integration
features <- SelectIntegrationFeatures(object.list = sample.list)

## Find anchors for the integration
condition.anchors <- FindIntegrationAnchors(object.list = sample.list, anchor.features = features)

## Create the integrated dataset
octo_all_stages_integrated <- IntegrateData(anchorset = condition.anchors)
```

# Run a basic analysis on the integrated Seurat object

```{r basic_seurat, echo=FALSE}
## Set default assay to the integrated one
DefaultAssay(octo_all_stages_integrated) <- "integrated"

## Run the standard workflow for PCA clustering
octo_all_stages_integrated <- ScaleData(octo_all_stages_integrated, verbose = FALSE)
octo_all_stages_integrated <- RunPCA(octo_all_stages_integrated, npcs = 50, verbose = FALSE)

## Visualize PCA results
DimPlot(octo_all_stages_integrated, reduction = "pca", label = T)
```

# Transfer the Scanpy annotation to the Seurat object

```{r transfer_annotation, echo=FALSE}
## Load barcodes from Scanpy
cell_metadata <- read.csv("all_stages_corrected_anno.csv")
cell_metadata$stage <- sub("_.*", "", cell_metadata$batch)
print(head(cell_metadata))
barcodes <- cell_metadata$barcode

## Add annotation of Scanpy to integrated dataset
target <- as.vector(octo_all_stages_integrated@assays[["integrated"]]@data@Dimnames[[2]])
cell_metadata_reversed <- as.data.frame(t(cell_metadata))
barcodes <- cell_metadata_reversed[1,]
colnames(cell_metadata_reversed) <- c(cell_metadata_reversed[1,])
t <- select(cell_metadata_reversed, which(names(cell_metadata_reversed) %in% target))
cell_metadata_rearranged <- as.data.frame(t(t)) %>%
  arrange(factor(barcode, levels = c(octo_all_stages_integrated@assays[["integrated"]]@data@Dimnames[[2]])))
octo_all_stages_integrated[['Scanpy_annot']] <-cell_metadata_rearranged$anno[match(rownames(octo_all_stages_integrated@meta.data), cell_metadata_rearranged$barcode)]

## Add embryonic stage information in metadata of Seurat object
octo_all_stages_integrated[['Stage']] <-cell_metadata_rearranged$stage[match(rownames(octo_all_stages_integrated@meta.data), cell_metadata_rearranged$barcode)]
```

# Run UMAP clustering of Seurat

```{r umap_seurat, echo=FALSE}
## Look at the Elbow plot to determine best number of dimensions
ElbowPlot(octo_all_stages_integrated, ndims = 50)

## Search best resolution using clustree
octo_all_stages_integrated <- FindNeighbors(octo_all_stages_integrated, dims = 1:50, reduction = "pca")
octo_all_stages_integrated <- FindClusters(octo_all_stages_integrated, resolution = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1, 1.1, 1.2, 1.3, 1.4, 1.5))
clustree(octo_all_stages_integrated)

## Set best resolution as default for the Seurat object
octo_all_stages_integrated <- FindClusters(octo_all_stages_integrated, resolution = 0.7)

## Run the UMAP clustering
octo_all_stages_integrated <- RunUMAP(octo_all_stages_integrated, dims = 1:35, reduction = "pca", reduction.name = "umap.integrated")

## Visualize clustering with both Sanpy annotations and Stages
DimPlot(octo_all_stages_integrated, reduction = "umap.integrated", group.by = c("Scanpy_annot"), label=TRUE)
DimPlot(octo_all_stages_integrated, reduction = "umap.integrated", group.by = c("Stage"), label=TRUE)

## Save the Seurat object as a RDS object
saveRDS(octo_all_stages_integrated, file = "octo_all_stages_integrated.rds")
```

# Downsample GLIA cells

```{r downsample_GLIA, echo=FALSE}
## Load saved Seurat object
octo_all_stages_integrated <- readRDS("octo_all_stages_integrated.rds")

## Subselect the cells annotated as GLIA in Scanpy annotation
octo_glia <- subset(octo_all_stages_integrated, subset = Scanpy_annot == "GLIA")

## Visualize the GLIA cells
DimPlot(octo_glia, reduction = "umap.integrated", group.by = c("Scanpy_annot"), label=TRUE)

## Run Seurat UMAP clustering on these cells
octo_glia <- ScaleData(octo_glia, verbose = FALSE)
octo_glia <- RunPCA(octo_glia, npcs = 50, verbose = FALSE)
ElbowPlot(octo_glia, ndims = 50)
octo_glia <- FindNeighbors(octo_glia, dims = 1:35, reduction = "pca")
octo_glia <- FindClusters(octo_glia, resolution = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1, 1.1, 1.2, 1.3, 1.4, 1.5))
clustree(octo_glia)
octo_glia <- FindClusters(octo_glia, resolution = 0.8)
octo_glia <- RunUMAP(octo_glia, dims = 1:35, reduction = "pca", reduction.name = "umap.integrated")

## Visualize clustering of the GLIA cells with both Sanpy annotations and Stages
DimPlot(octo_glia, reduction = "umap.integrated", group.by = c("Scanpy_annot"), label=TRUE)
DimPlot(octo_glia, reduction = "umap.integrated", group.by = c("Stage"), label=TRUE)

## Save this new Seurat object as a RDS object
saveRDS(octo_glia, file = "octo_glia.rds")
```

# Convert GLIA Seurat object to Cell Data Set (CDS) object for Monocle

```{r conversion, echo=FALSE, fig.height=7, fig.width=10}
## Load GLIA seurat object
octo_glia <- readRDS("octo_glia.rds")

## Set default assay to RNA and join the different layers for the conversion
DefaultAssay(octo_glia) <- "RNA"
octo_glia <- JoinLayers(octo_glia)

## Convert the Seurat object to a CDS object
octo_glia_CDS <- as.cell_data_set(octo_glia)

## Add important information that was not transferred
rowData(octo_glia_CDS)$gene_short_name <- rownames(octo_glia_CDS)
octo_glia_CDS <- estimate_size_factors(octo_glia_CDS)

## Preprocess and do first tSNE clustering with Monocle
octo_glia_CDS <- preprocess_cds(octo_glia_CDS, num_dim = 60)
plot_pc_variance_explained(octo_glia_CDS)
octo_glia_CDS <- reduce_dimension(octo_glia_CDS, reduction_method = "tSNE")

## Visualize this new clustering with the 3 embryonic stages
plot_cells(octo_glia_CDS, reduction_method="tSNE", color_cells_by = "Stage")
```

# Remove the batch effect from the CDS object

```{r remove_batch_effect, echo=FALSE, fig.height=7, fig.width=10}
## Visualize this new clustering with the different batches
plot_cells(octo_glia_CDS, reduction_method = "tSNE", color_cells_by = "orig.ident", label_cell_groups = FALSE)

## Remove the batch effect
octo_glia_CDS <- align_cds(octo_glia_CDS, num_dim = 60, alignment_group = "orig.ident")

## Run a new clustering on the corrected dataset
octo_glia_CDS <- reduce_dimension(octo_glia_CDS, reduction_method = "tSNE")
reducedDims(octo_glia_CDS)$UMAP <- reducedDims(octo_glia_CDS)$tSNE

## Visualize this new clustering with the different batches and the Stages to see if the batch correction worked
plot_cells(octo_glia_CDS, reduction_method="UMAP", color_cells_by = "orig.ident", label_cell_groups = FALSE)
plot_cells(octo_glia_CDS, reduction_method="UMAP", color_cells_by = "Stage", label_cell_groups = FALSE)
```

# Improve the clustering by optimizing the resolution

```{r improve_clustering, echo=FALSE, fig.height=8, fig.width=10}
## Searched best clustering resolution using silhouette scores
### Create a dataframe of the different resolutions to try
resolutions <- c(1e-3, 1.5e-3, 2e-3, 2.5e-3, 3e-3, 3.5e-3, 4e-3, 4.5e-3, 5e-3, 5.5e-3, 6e-3, 6.5e-3, 7e-3, 7.5e-3, 8e-3, 8.5e-3, 9e-3, 9.5e-3, 1e-2)
silhouette_scores <- data.frame(res = numeric(), mean_silhouette = numeric())

### Create a matrix containing the UMAP coordinates
umap_coords <- reducedDims(octo_glia_CDS)[["UMAP"]]
dist_matrix <- dist(umap_coords)

### Loop through the different resolutions to calculate the silhouette scores
for (res in resolutions) {
  octo_glia_CDS <- cluster_cells(octo_glia_CDS, resolution = res)
  curr_clusters <- as.numeric(clusters(octo_glia_CDS))
  if (length(unique(curr_clusters)) > 1) {
    sil <- silhouette(curr_clusters, dist_matrix)
    mean_sil <- mean(sil[, 3])
    silhouette_scores <- rbind(silhouette_scores, data.frame(res = res, mean_silhouette = mean_sil))
  }
}

### Visualize the silhouette scores
ggplot(silhouette_scores, aes(x = as.factor(res), y = mean_silhouette, group = 1)) +
  geom_line(color = "steelblue") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Silhouette Analysis by Resolution",
       x = "Resolution (Log)", y = "Average Silhouette score")

## Searched best clustering resolution using clustree
### Run clustering at multiple resolutions
res_range <- seq(1e-3, 9.5e-3, length.out = 18)
octo_glia_CDS_list <- lapply(res_range, function(r) {
    cluster_cells(octo_glia_CDS, resolution = r)
})

### Extract cluster assignments into a data frame
cluster_matrix <- sapply(octo_glia_CDS_list, function(x) clusters(x))
colnames(cluster_matrix) <- paste0("res_", res_range)

### Run clustree and visualize the 
clustree(cluster_matrix, prefix = "res_")

# Cluster the cells with the best resolution from the 2 methods 
octo_glia_CDS <- cluster_cells(octo_glia_CDS, reduction_method="UMAP", resolution=7e-3)
```

# Calculate trajectories of this clustering and visualize them

```{r calculate_trajectory, echo=FALSE, fig.height=7, fig.width=10}
## Calculate trajectories
octo_glia_CDS <- learn_graph(octo_glia_CDS)

## Visualize the trajectories
plot_cells(octo_glia_CDS, label_cell_groups = FALSE)
```

# Create a pdf file of the trajectory plot

```{r trajectory_pdf, echo=FALSE, fig.height=7, fig.width=10}
## Set colors of the groups
my_stage_colors <- c(
  "IX" = "#CC5500",
  "XII" = "#6082B6",
  "XV" = "#8A9A5B"
)

## Create the plot
p <- plot_cells(octo_glia_CDS, color_cells_by = "Stage", label_cell_groups = FALSE, cell_size = 1) +
  scale_color_manual(values = my_stage_colors)
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$colour <- NA
p1 <- p + scale_color_manual(values = my_stage_colors)

## Save it as a pdf
ggsave(filename = "Trajectory_plot_GLIA.pdf", 
       plot = p1, 
       device = "pdf", 
       width = 10,
       height = 8,
       useDingbats = FALSE)
```

# Define the root node of the trajectory and visualize the result

```{r root_node, echo=FALSE}
## Set the function to define automatically the best root node based on the embryonic stage
get_earliest_principal_node <- function(cds, time_bin="IX"){
  cell_ids <- which(colData(cds)[, "Stage"] == time_bin)
  
  closest_vertex <-
  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <-
  igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
  (which.max(table(closest_vertex[cell_ids,]))))]
  
  root_pr_nodes
}

## Set the root node
octo_glia_CDS <- order_cells(octo_glia_CDS, root_pr_nodes=get_earliest_principal_node(octo_glia_CDS))

## Save the CDS object as a RDS file
saveRDS(octo_glia_CDS, file = "octo_glia_CDS.rds")
```

# Create a pdf file of the pseudotime plot

```{r pseudotime_pdf, echo=FALSE}
## Create the plot
p <- plot_cells(octo_glia_CDS,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=TRUE,
           label_branch_points=TRUE,
           graph_label_size=1.5,
           cell_size = 1)
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$colour <- NA

## Save it as a pdf
ggsave(filename = "Trajectory_with_pseudotime_GLIA.pdf", 
       plot = p, 
       device = "pdf", 
       width = 10,
       height = 8,
       useDingbats = FALSE)
```

# Highlighting cells of each pattern from the NMF in the trajectory plot

```{r highlight_cells_pattern, echo=FALSE, fig.width=16, fig.height=12}
## Read table file containing enrichment values for the different patterns for all the cell barcodes
barcodes_pattern <- read.csv("patterns_barcodes_for_monocle.csv")

## Reformat cell barcodes names
barcodes_pattern$barcode <- sub("^(.*)(-1-)(.*)$", "\\3_\\1-1", barcodes_pattern$barcode)

## Define function to rescale the enrichment values to percentage of enrichment
rescale_percent <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

## Create a list to store the plots of each pattern
plot_list_percent <- list()

## Define the function to create the plot for each pattern
for(p_col in paste0("P_", 1:9)) {
  
  ### Create a vector of scores
  all_cells <- colnames(octo_glia_CDS)
  scores <- setNames(rep(min(barcodes_pattern[[p_col]]), length(all_cells)), all_cells)
  present_barcodes <- intersect(barcodes_pattern$barcode, all_cells)
  scores[present_barcodes] <- barcodes_pattern[[p_col]][match(present_barcodes, barcodes_pattern$barcode)]
  
  ### Convert to percentages
  scores_percent <- rescale_percent(scores)
  
  ### Add the scores to the CDS object
  colData(octo_glia_CDS)[[p_col]] <- scores_percent
  
  ### Generate the plot
  p <- plot_cells(octo_glia_CDS,
                  color_cells_by = p_col,
                  show_trajectory_graph = TRUE,
                  label_cell_groups = FALSE,
                  label_leaves = FALSE,
                  cell_size = 0.8,
                  cell_stroke = 0) +
       scale_color_gradientn(colors = c("lightgrey", "yellow", "red"), 
                             labels = scales::percent,
                             name = "Intensity") +
       ggtitle(paste("Gradient of expression :", p_col)) +
    guides(color = guide_colorbar(order = 1)) +
    theme(panel.background = element_blank())
  p$layers[[1]]$aes_params$stroke <- 0
  p$layers[[1]]$aes_params$colour <- NA
  
  plot_list_percent[[p_col]] <- p
}
```

# Create patchwork plot highlighting the cells enriched for each NMF pattern and save it as a pdf

```{r plot_highlighted_cells_pattern, echo=FALSE, fig.width=16, fig.height=12}
wrap_plots(plot_list_percent, ncol = 3)
p <- wrap_plots(plot_list_percent, ncol = 3)
ggsave(filename = "Trajectory_gradient_pattern_GLIA.pdf", 
       plot = p, 
       device = "pdf", 
       width = 16,
       height = 12,
       useDingbats = FALSE)
```

# Plot Pseudotimes against patterns to see if they correlate with the results of the NMF analysis

```{r pseudotime_vs_patterns, echo=FALSE, fig.width=10, fig.height=6}
## Retrieve separated scores of each pattern from table file of previous step
pattern_cols <- grep("^P_", colnames(barcodes_pattern), value = TRUE)
scores <- barcodes_pattern[, pattern_cols]

## Scaling to turn all values to scores between 0 and 1
scores_pos <- apply(scores, 2, function(x) (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE)))

## Conversion to percentages
df_pct <- as.data.frame(t(apply(scores_pos, 1, function(x) x / sum(x))))

## Re-integration of the cell barcodes
df_pct$barcode <- barcodes_pattern$barcode
pseudo_df <- data.frame(
  barcode = row.names(colData(octo_glia_CDS)),
  pseudotime = pseudotime(octo_glia_CDS)
)

## Fusion of percentages and pseudotimes
final_df <- merge(df_pct, pseudo_df, by = "barcode")

## Create a summary table of this fusion
summary_patterns <- final_df %>%
  ## Convert to long format
  pivot_longer(cols = starts_with("P_"), names_to = "Pattern", values_to = "Enrichment") %>%
  ## Cleaning: removing NAs and infinite pseudotimes
  filter(!is.na(pseudotime), !is.na(Enrichment), is.finite(pseudotime)) %>%
  group_by(Pattern) %>%
  summarise(
    mean_pseudo = sum(pseudotime * Enrichment, na.rm = TRUE) / sum(Enrichment, na.rm = TRUE),
    sd_pseudo = sd(pseudotime, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  ## Sorting by pattern (P_1, P_2, etc.)
  mutate(Pattern = factor(Pattern, levels = paste0("P_", 1:9))) %>%
  arrange(Pattern)

## Check result
print(summary_patterns)

## Visualize the result as a curve
ggplot(summary_patterns, aes(x = Pattern, y = mean_pseudo, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  theme_minimal() +
  theme(axis.title.y = element_text(size = 18, face = "bold", margin = margin(r = 10)),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14)) + 
  labs(title = "", x = "", y = "Average pseudotime")

## Save the plot as a pdf
p <- ggplot(summary_patterns, aes(x = Pattern, y = mean_pseudo, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  theme_minimal() +
  theme(axis.title.y = element_text(size = 18, face = "bold", margin = margin(r = 10)),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14)) + 
  labs(title = "", x = "", y = "Average pseudotime")
ggsave(filename = "Pseudotime_vs_patterns.pdf", 
       plot = p, 
       device = "pdf", 
       width = 10,
       height = 6,
       useDingbats = FALSE)
```